\Huge{Vipassana for Hackers} \\ \large{Paper Two: The Brain} \\ \large\textit{Version 0.2}

### 2019-09-30 paper-two/paper-two.tex

@@

There have been studies
that have shown that once you bring your attention back to breath, the neural
structures involved in the control of the Autonomic Nervous System and attention
start firing more actively (in this case, the DL-PFC --- a part of the CEN and the
Hippocampus and ACC --- both part of the SN). There are also global dampening changes
seen in the brain, particularly within the DMN and the Amygdala, which is involved in
the flight-or-fight (stress) response. These changes have been
termed the ``relaxation response'', as they are antagonistic to the stress
response. This relaxation response can be thought of as a gateway to altered states
of mind. \cite{relaxationresponse}

@@

This is hard at first, with the body involuntarily reacting by jerking and writhing
in involuntary response to the act of observation. Initially, your awareness is
itself jerky and it is hard to observe sensations consciously. There is also the
pain. This eclipses all other sensations. In my case, it was difficult to maintain
awareness of any other sensation with the pain that was emanating from my legs and
back. The OFC is involved in assigning emotion to sensation \cite{ofcemotion} and the
ACC plays a vital role in perception of sensation. These two parts of the brain are
usually functionally connected. In the beginning, as I sat trying to observe
sensations but instead was only aware of the pain, the ACC was firing rapidly and the
OFC then assigned these signals to the Limbic System (amygdala and
hippocampus)... which decided to put me in a less than ideal emotional state.

But as your awareness gets smoother and observations get more objective, this
functional connectivity between the ACC and the OFC slowly starts changing. As you
continue to pay attention to the different sensations you attempt to avoid assigning
emotions to them. Without assigning an emotional response, you can observe with a
non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
but the Limbic System is quiesced and, in turn, you react less. It is as if both your
conscious and unconscious awarenesses are slowly aligning until you experience some
occasions when it feels as if there is nothing but pure, detached awareness moving
through your body (or some parts of your body). Senses are not dulled, however. In
addition to a feeling of emotional detachment, the Insula and the Somatosensory
Cortex, which are involved in sensory awareness, were firing \textbf{more} actively,
not less.

(The somatosensory cortex is located in the gyrus and the parietal lobes:) \todo{Maybe move sensory
  homunculus diagram and explanation to the anatomy section?}

\todo{add sensory homunculus diagram}
Somatosensory and Motor cortices

Perhaps the strangest experience during a Vipassana course is that of shifts in your
sense of self. The ``self'', as a subjective agent in space and time changes, if at
first only very slightly. As you close your eyes and meditate, it may feel as though
the body has shifted in space or rotated in different directions. During these times,
the vmPFC and PCC become less active. As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the Insula
fires more rapidly, leading to alternative networks of consciousness arising. It is
as if losing the influence of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain halucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

Vipassana quiets the DMN, causes certain parts of the SN and CEN to fire more
rapidly. That is, reacting parts of the network and portions of the network which are
aware of bodily sensation, respectively. This causes changes to a person's behavior
(in the short term, at least) and these changes can often be seen as soon as someone
leaves the course on the tenth day.

But repeated practice is essential for any sustained neuroplasticity since rewriting
many years of habit formation requires more than a ten day course. However, the ten
day course is well-designed and will give someone experimenting with meditation the
minimum observational change neccesary to instill an interest, to keep practicing.

There have been studies looking at both the structural changes in the brain and
functional changes in brain activity, especially to the DMN and the SN. But such
studies are usually only done on meditators who have clocked thousands of hours of
practice, which diverges from the purpose of this paper. \todo{add references with
  evidence of structural changes in the brain after meditation}

How is Vipassana different from simply concentrating on a task, as hard as one can?
When concentrating on any task, your DMN is quieted, the SN and CEN are firing, and you
enter into a state of flow. But flow or no flow, the reacting part of the brain is
also working. And it is working hard. The goal of Vipassana is not simply to
concentrate or two enter a state of flow, but to break old reactionary habit patterns
altogether.

@@

\bibitem{cortexbliss}
  Picard, Fabienne.
  \textit{State of belief, subjective certainty and bliss as a product of cortical
    dysfunction}
  Cortex, Volume 49, Issue 9, Pages 2494–2500ISSN 0010 9452,
  https://doi.org/10.1016/j.cortex.2013.01.006. 2013.

>>>>

The following are my observations and speculations about what could be happening 
in the brain during a 10 course.

@@

There have been studies that have shown that once you bring your 
attention back to breath, the neural structures involved in the control
of the Autonomic Nervous System and attention start firing more actively
(in this case, the DL-PFC --- a part of the CEN and the Hippocampus a part of the DMN 
and ACC --- a part of the SN). There are also global dampening changes seen in the brain, 
particularly within the DMN and the Amygdala, which is involved in
the flight-or-fight (stress) response. These changes have been
termed the ``relaxation response'', as they are antagonistic to the stress
response. This relaxation response can be thought of as a gateway to altered states
of mind. \cite{relaxationresponse}

@@

This is hard at first, with the body involuntarily reacting by jerking and writhing
in involuntary response to the act of observation. Initially, your awareness is
itself jerky and it is hard to observe sensations consciously. There is also the
pain. This eclipses all other sensations. In my case, it was difficult to maintain
awareness of any other sensation with the pain that was emanating from my legs and
back. The OFC is involved in assigning emotion to sensation \cite{ofcemotion} and the
ACC plays a vital role in perception of sensation. These two parts of the brain are
usually functionally connected. In the beginning, as I sat trying to observe
sensations and was only aware of the pain, the ACC was firing rapidly and the
OFC then assigned these signals to the Limbic System which 
put me in a less than ideal emotional state.

But as your awareness gets smoother and observations get more objective, this
functional connectivity between the ACC and the OFC slowly starts changing. As you
continue to pay attention to the different sensations and you attempt to avoid assigning
emotions to them. Without assigning an emotional response, you can observe with a
non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
but the Limbic System is quiesced and, in turn, you react less. It is as if both your
conscious and unconscious awarenesses are slowly aligning until you experience some
occasions when it feels as if there is nothing but detached awareness moving
through your body (or some parts of your body). Senses are not dulled, however. In
addition to a feeling of emotional detachment, the Insula and the Somatosensory
Cortex, which are involved in sensory awareness, were firing \textbf{more} actively,
not less.

Perhaps the strangest experience during a Vipassana course is that of shifts in your
sense of self. The ``self'', as a subjective agent in space and time changes, if at
first only very slightly. As you close your eyes and meditate, it may feel as though
the body has shifted in space or rotated in different directions. During these times,
the vmPFC and PCC become less active. As the meditations get deeper the DMN
occasionally quiets completely (perhaps only for a short duration) while the Insula
fires more rapidly, leading to alternative networks of consciousness arising. It is
as if losing the influence of one's own narrative leads to new insights that are
usually kept from consciousness. It has been found that experiencing this, even for a
few seconds, can lead to lasting changes long after the activity of meditation itself
has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
seizures where the abnormal activity of the Anterior Insular Cortex leads to
heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}

@@

Vipassana causes global dampening changes in the DMN, causes certain parts of 
the SN and CEN to fire more rapidly. That is, reacting parts of the network and 
portions of the network which are aware of bodily sensation, respectively. 
This changes a person's behavior (in the short term, at least) and these 
changes can often be seen as soon as someone leaves the course on the tenth day.

But repeated practice is essential for any sustained neuroplasticity since rewriting
many years of habit formation requires more than a ten day course. However, the ten
day course is well-designed and will give someone experimenting with meditation the
minimum observational change necessary to instill an interest, to keep practicing.

There have been studies looking at both the structural changes in the brain and
functional changes in brain activity, especially to the DMN and the SN. But such
studies are usually only done on meditators who have clocked thousands of hours of
practice, which diverges from the purpose of this paper. \cite {tibeteanmonks
, experiencemed, experiencemeda}

How is Vipassana different from simply concentrating on a task, as hard as one can?
When concentrating on any task, your DMN is quieted, the SN and CEN are firing, and you
enter into a state of flow. But flow or no flow, the reacting part of the brain is
also working. And it is working hard. The goal of Vipassana is not simply to
concentrate or to enter a state of flow, but to break old reactionary habit patterns
altogether.

@@

\bibitem{cortexbliss}
  Picard, Fabienne.
  \textit{State of belief, subjective certainty and bliss as a product of cortical
           dysfunction}
   Cortex, Volume 49, Issue 9, Pages 2494–2500ISSN 0010 9452,
   \url: https://doi.org/10.1016/j.cortex.2013.01.006. 2013.

\bibitem{experiencemed}
   Brewer JA, Worhunsky PD, Gray JR, Tang YY, Weber J, Kober H.
   \textit {Meditation experience is associated with differences in default mode 
          network activity and connectivity.}
     Proc Natl Acad Sci U S A. 2011 Dec 13;108(50):20254-9. 
     doi: 10.1073/pnas.1112029108.

\bibitem{experiencemeda}
  Zoran Josipovic, Ilan Dinstein, Jochen Weber, David J. Heeger
  \textit {Influence of meditation on anti-correlated networks in the brain}
   Front. Hum. Neurosci., 03 January 2012,
   https://doi.org/10.3389/fnhum.2011.00183
